Staged Abstract Interpreters

### 2019-09-16 paper/10_intro.tex
misc

To apply the Futamura projections to abstract interpreters, we consider
\citet{DBLP:journals/pacmpl/DaraisLNH17}'s big-step monadic abstract interpreter
as the unstaged baseline.  Similar to two-level semantics \cite{NIELSON1989117},
monads provide a good abstraction to hide the detail of abstract interpretation
semantics. However, monads do not have a clear stage distinction. We further
introduce binding-times and make the monads binding-time polymorphic. Our
proposed framework adopts type-based multi-stage programming from the
Lightweight Modular Staging framework, and implements the first Futamura projection of
an abstract interpreter for a stateful higher-order language. By deriving staged
monads that can be used to perform code generation, we obtain a staged abstract
interpreter that can be used for specialization. Through staging, the generated
code is specialized to the input program, and all the monadic operations are
inlined and compiled down to low-level Scala code.

@@

\paragraph{Applications and Evaluation}
We evaluate the idea of staging an abstract interpreter through
case studies and an empirical performance evaluation.
1) We compare our approach with abstract compilation
\cite{Boucher:1996:ACN:647473.727587}, an implementation technique for
control-flow analyses. We show that by utilizing type-based stage
annotations, we can achieve the same optimizations. Meanwhile,
the analyzer does not need to change (modulo the addition of stage annotations),
thereby requiring significantly less engineering effort.
2) We extend the basic staged abstract interpreter to cover different flow
analyses, including a store-widened analysis, a context-sensitive
analysis, and abstract garbage collection.
3) We show that staging an abstract interpreter enables modular
compilation of an analysis to programs. Here we borrow the concept of
modular analysis, and show that the compiled analysis is reusable.
Therefore, the approach provides a modular way to create optimized
analysis code by mechanized reuse of a whole-program analyzer.
4) We empirically evaluate the performance improvements gained by staging,
showing an order of magnitude speedup.

@@

\paragraph{Organization} The paper is organized as follows:
\begin{itemize}[leftmargin=2em]
  \item We begin by introducing our target language and reviewing
    monads in Scala, and then presenting the generic interpreter
    (Section~\ref{prelim}), after which we review instantiations
    of concrete interpretation (Section~\ref{unstaged_conc}) and
    staged concrete interpretation (Section~\ref{stagedinterp}).
  \item We present the unstaged abstract interpreter under the same
    framework by replacing the environment, store, and values with their
    abstract counterparts (Section~\ref{unstaged_abs}). We then
    show that the combination of approximation and specialization, dubbed
    \textit{staged abstract interpreters}, can be readily derived
    (Section~\ref{sai}). We also summarize the approach and discuss
    soundness properties after showing the four artifacts.
  \item We summarize our approach and discuss correctness (Section~\ref{discussion}).
    We conduct three case studies (Section~\ref{cases_study}) and
    an empirical evaluation on performance improvements (Section~\ref{evaluation}).
    Finally we discuss related work and conclude.
\end{itemize}

>>>>

To apply the Futamura projections to abstract interpreters, we consider
\citet{DBLP:journals/pacmpl/DaraisLNH17}'s big-step monadic abstract
interpreter as the unstaged baseline.  Our proposed framework adopts type-based
MSP from the Lightweight Modular Staging framework, and implements the first
Futamura projection of this monadic abstract interpreter for a stateful
higher-order language. 
Similar to two-level semantics \cite{NIELSON1989117}, monads provide a good
abstraction to hide the detail of abstract interpretation semantics. However,
monads do not have a clear stage distinction. We further introduce
binding-times and make the monads binding-time polymorphic. 
By deriving staged monads that can be used to perform code generation, we
obtain a staged abstract interpreter that can be used for specialization.
Through staging, the generated code is specialized to the input program, and
all the monadic operations are inlined and compiled down to low-level Scala
code.

@@

\paragraph{Applications and Evaluation}
We evaluate the idea of staging an abstract interpreter through
case studies and an empirical performance evaluation.
1) We compare our approach with abstract compilation
\cite{Boucher:1996:ACN:647473.727587}, an implementation technique for
control-flow analyses. We show that by utilizing type-based stage
annotations, we can achieve the same optimizations. Meanwhile,
the analyzer does not need to change (modulo the addition of stage annotations),
thereby requiring significantly less engineering effort.
2) We extend the basic staged abstract interpreter to cover different flow
analyses, including a store-widened analysis, a context-sensitive
analysis, and abstract garbage collection.
3) We show that staging an abstract interpreter enables modular
compilation of an analysis to programs. Here, we borrow the concept of
modular analysis, and show that the compiled analysis is reusable.
Therefore, the approach provides a modular way to create optimized
analysis code by mechanized reuse of a whole-program analyzer.
4) We empirically evaluate the performance improvements gained by staging,
showing an order of magnitude speedup.

@@

\paragraph{Organization} The paper is organized as follows:
\begin{itemize}[leftmargin=2em]
  \item We begin by introducing our target language and reviewing
    monads in Scala, and then presenting the generic interpreter
    (Section~\ref{prelim}), after which we review instantiations
    of concrete interpretation (Section~\ref{unstaged_conc}) and
    staged concrete interpretation (Section~\ref{stagedinterp}).
  \item We present the unstaged abstract interpreter under the same
    framework by replacing the environment, store, and values with their
    abstract counterparts (Section~\ref{unstaged_abs}). We then
    show that the combination of approximation and specialization, dubbed
    \textit{staged abstract interpreters}, can be readily derived
    (Section~\ref{sai}). We also summarize the approach and discuss
    correctness properties after showing the four artifacts (Section~\ref{discussion}).
  \item We conduct three case studies (Section~\ref{cases_study}) and an
    empirical evaluation on performance improvements (Section~\ref{evaluation}).  
    Finally we discuss related work and conclude.
\end{itemize}

### 2019-09-16 paper/20_prelim.tex
misc

The abstract syntax we present can be seen as a deep embedding of the language:
we use data-types to represent programs. This is the most natural choice
for program analysis and allows us to use different interpretations over the
AST; with the inheritance and overriding mechanisms in Scala, we may also add
new language constructs and reuse existing interpretations.

@@

Note that here the monadic object of type @List[_]@ encapsulates the data internally.
Therefore, it only exposes the simplified version of @flatMap@, where the monadic
value of @M[A]@ is not introduced as a function argument. The trait @MonadOps@ (Figure
\ref{fig:monadops}) defines the simplified versions of monadic operations that
are necessary for @for@-comprehension. The conversion between @Monad@ and
@MonadOps@ can be achieved by using the implicit design pattern.
In the rest of the paper, we use Scala's @for@-comprehension syntax and a few
monads and monad transformers such as @ReaderT@, @StateT@, and @SetT@ to
write our interpreters.
Monad transformers are type constructors of kind @(* --> *) --> (* --> *)@, which
take a monad type as argument and produce another monad type. By using monad
transformers, we can combine multiple monads into a single one.  The
implementation of monads and monad transformers follows the ones from Haskell.

@@

In this section, we present the generic evaluator interface in the style of a big-step
definitional interpreter. The key idea is to keep both the binding-time type and
returned monadic type abstract, so that they can be instantiated
differently. We also need to abstract the primitive operations on those types.
In later sections, we will instantiate the monadic type to perform concrete
interpretation \cite{DBLP:conf/popl/LiangHJ95} or abstract interpretation
\cite{Sergey:2013:MAI:2491956.2491979, DBLP:journals/pacmpl/DaraisLNH17}.

>>>>

The abstract syntax we present can be seen as a deep embedding of the language:
we use data-types to represent programs. This is the most natural choice
for program analysis and it allows us to use different interpretations over the
AST; with the inheritance and overriding mechanisms in Scala, we may also add
new language constructs and reuse existing interpretations.

@@

Note that here the monadic object of type @List[_]@ encapsulates the data internally.
Therefore, it exposes the simplified interface of @flatMap@, where the monadic
value of @M[A]@ is not introduced as a function argument. The trait @MonadOps@ (Figure
\ref{fig:monadops}) defines the simplified signatures of monadic operations that
are necessary for @for@-comprehension. The conversion between @Monad@ and
@MonadOps@ can be achieved by using the implicit design pattern.
In the rest of the paper, we use Scala's @for@-comprehension syntax and a few
monads and monad transformers such as @ReaderT@, @StateT@, and @SetT@ to
write our interpreters.
Monad transformers are type constructors of kind @(* --> *) --> (* --> *)@, which
take a monad type as argument and produce another monad type. By using monad
transformers, we can combine multiple monads into a single one.  The
implementation of monads and monad transformers follows the ones from Haskell.

@@

In this section, we present the generic evaluator interface in the style of a big-step
definitional interpreter. The key idea is to keep both the binding-time type and
returned monadic type abstract, so that they can be instantiated
differently. We also need to abstract the primitive operations on those types.
In later sections, we will instantiate the monadic type to perform concrete
interpretation \cite{DBLP:conf/popl/LiangHJ95, Wadler:1992:EFP:143165.143169}
or abstract interpretation \cite{Sergey:2013:MAI:2491956.2491979,
DBLP:journals/pacmpl/DaraisLNH17}.
